---
title: "Multi-Environment Trials with speed"
author: "Wasin Pipattungsakul"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Multi-Environment Trials with speed}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# MET Design

## Overview

Multi-environment trials (MET) are experimental designs used to evaluate the performance of treatments across
different environments. An environment typically represents a unique combination of location, year, season, or
management practice. MET designs are essential for understanding genotype × environment (G×E) interactions,
stability, and adaptability of treatments under different conditions.

```{r load-package}
library(speed)
```

## When to Use

- Treatment evaluation across multiple locations and/or years
- Crop breeding and regional recommendation trials
- Testing stability and adaptability
- When results are intended for large geographic areas

## Structure

- **Sites**: Different locations of the trial
- **Site-blocks**: Blocks within sites

## Optimising Allocation across All Sites

### Setting Up MET Design with speed

Now we can create a data frame representing a MET design. Note that we can specify different dimensions for each
site in the `designs` argument.

```{r met-df}
all_treatments <- c(rep(1:50, 7), rep(51:57, 8))
met_design <- initialise_design_df(
  items = all_treatments,
  designs = list(
    a = list(nrows = 28, ncols = 5, block_nrows = 7, block_ncols = 5),
    b = list(nrows = 20, ncols = 3, block_nrows = 10, block_ncols = 3),
    c = list(nrows = 20, ncols = 4, block_nrows = 10, block_ncols = 4),
    d = list(nrows = 15, ncols = 4, block_nrows = 5, block_ncols = 4),
    e = list(nrows = 22, ncols = 3, block_nrows = 11, block_ncols = 3)
  )
)

met_design$site_col <- paste(met_design$site, met_design$col, sep = "_")
met_design$site_block <- paste(met_design$site, met_design$block, sep = "_")

head(met_design)
```

```{r met-plot1}
#| echo: false
met_design$block <- factor(met_design$block)

ggplot2::ggplot(met_design, ggplot2::aes(col, row, fill = block)) +
  ggplot2::geom_tile(color = "black") +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::facet_wrap(~site, scales = "free") +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = 1:max(met_design$col)) +
  ggplot2::scale_y_continuous(expand = c(0, 0), breaks = 1:max(met_design$row), trans = scales::reverse_trans()) +
  ggplot2::theme_bw() +
  ggplot2::theme(panel.grid.major = ggplot2::element_blank(), panel.grid.minor = ggplot2::element_blank())
```

### Performing the Optimisation

For MET designs, we use lists of named arguments to specify the hierarchical structure. The `optimise` parameter
defines what to optimise and constraints at each level.

```{r met-example}
optimise <- list(
  connectivity = list(spatial_factors = ~site),
  balance = list(swap_within = "site", spatial_factors = ~ site_col + site_block)
)

met_result <- speed(
  data = met_design,
  swap = "treatment",
  early_stop_iterations = 5000,
  optimise = optimise,
  optimise_params = optim_params(random_initialisation = TRUE, adj_weight = 0),
  seed = 112
)

met_result
```

### Output of the Optimisation

The output shows optimisation results for the design. The score and iterations are combined for the entire
design, while the treatments, and stopping criteria are reported separately for each level, allowing you to
assess the quality of optimisation at each hierarchy level.

```{r met-output}
str(met_result)
```

No duplicated treatments along any row or column.

```{r met-no-dupes}
df <- met_result$design_df
max(table(df$treatment, df$site_col))
max(table(df$treatment, paste0(df$site, df$row)))
```

### Visualise the Output

```{r met-plot2}
ggplot2::ggplot(met_result$design_df, ggplot2::aes(col, row, fill = treatment)) +
  ggplot2::geom_tile(color = "black") +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::facet_wrap(~site, scales = "free") +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = 1:max(met_design$col)) +
  ggplot2::scale_y_continuous(expand = c(0, 0), breaks = 1:max(met_design$row), trans = scales::reverse_trans()) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )
```

This design has now been optimised at both the connectivity between sites and the balance within each site.

## Optimising Allocation across Some Sites

### Setting Up MET Design with speed

Now we can create a data frame representing a MET design. Note that we can specify different dimensions for each
site in the `designs` argument.

```{r met2-df}
fixed_treatments <- rep(1:54, 3)
non_fixed_treatments <- c(rep(1:50, 5), rep(51:54, 4))
met_design <- initialise_design_df(
  designs = list(
    a = list(items = fixed_treatments, nrows = 27, ncols = 6, block_nrows = 9, block_ncols = 6),
    b = list(items = non_fixed_treatments[1:60], nrows = 20, ncols = 3, block_nrows = 10, block_ncols = 3),
    c = list(items = non_fixed_treatments[1:80 + 60], nrows = 20, ncols = 4, block_nrows = 10, block_ncols = 4),
    d = list(items = non_fixed_treatments[1:60 + 140], nrows = 15, ncols = 4, block_nrows = 5, block_ncols = 4),
    e = list(items = non_fixed_treatments[1:66 + 200], nrows = 22, ncols = 3, block_nrows = 11, block_ncols = 3)
  )
)

met_design$site_col <- paste(met_design$site, met_design$col, sep = "_")
met_design$site_block <- paste(met_design$site, met_design$block, sep = "_")
met_design$allocation <- "free"
met_design$allocation[met_design$site == "a"] <- "fixed"

head(met_design)
```

```{r met2-plot1}
#| echo: false
met_design$block <- factor(met_design$block)

ggplot2::ggplot(met_design, ggplot2::aes(col, row, fill = block)) +
  ggplot2::geom_tile(color = "black") +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::facet_wrap(~site, scales = "free") +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = 1:max(met_design$col)) +
  ggplot2::scale_y_continuous(expand = c(0, 0), breaks = 1:max(met_design$row), trans = scales::reverse_trans()) +
  ggplot2::theme_bw() +
  ggplot2::theme(panel.grid.major = ggplot2::element_blank(), panel.grid.minor = ggplot2::element_blank())
```

### Performing the Optimisation

For MET designs, we use lists of named arguments to specify the hierarchical structure. The `optimise` parameter
defines what to optimise and constraints at each level.

```{r met2-example}
optimise <- list(
  connectivity = list(swap_within = "allocation", spatial_factors = ~site),
  balance = list(swap_within = "site", spatial_factors = ~ site_col + site_block)
)

met_result <- speed(
  data = met_design,
  swap = "treatment",
  early_stop_iterations = 5000,
  optimise = optimise,
  optimise_params = optim_params(random_initialisation = TRUE, adj_weight = 0),
  seed = 112
)

met_result
```

### Output of the Optimisation

The output shows optimisation results for the design. The score and iterations are combined for the entire
design, while the treatments, and stopping criteria are reported separately for each level, allowing you to
assess the quality of optimisation at each hierarchy level.

```{r met2-output}
str(met_result)
```

No duplicated treatments along any row or column.

```{r met2-no-dupes}
df <- met_result$design_df
max(table(df$treatment, df$site_col))
max(table(df$treatment, paste0(df$site, df$row)))
```

Site "a" maintains 3 replicates and no missing treatments in any other sites.

```{r met2-reps}
treatment_count <- table(df$treatment, df$site)
c(min(treatment_count[, "a"]), max(treatment_count[, "a"]))
c(min(treatment_count[, -1]), max(treatment_count[, -1]))
```

### Visualise the Output

```{r met2-plot2}
ggplot2::ggplot(met_result$design_df, ggplot2::aes(col, row, fill = treatment)) +
  ggplot2::geom_tile(color = "black") +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::facet_wrap(~site, scales = "free") +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = 1:max(met_design$col)) +
  ggplot2::scale_y_continuous(expand = c(0, 0), breaks = 1:max(met_design$row), trans = scales::reverse_trans()) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )
```

This design has now been optimised at both the connectivity between sites and the balance within each site.
# Spatial Design Considerations

## Field Shape and Orientation

## Neighbour Effects

# Using `speed` Effectively

1. **Set appropriate parameters**: Balance optimisation time with improvement <!-- Link to vignette on parameters -->
2. **[Visualise designs](autoplot.html)**: Always plot layouts before implementation  <!-- Link to autoplot -->
3. **Compare alternatives**: Test multiple blocking strategies
4. **Validate results**: Check constraint satisfaction and efficiency metrics

# Conclusion

## Further Reading

# Related Vignettes

*This vignette demonstrates the versatility of the `speed` package for agricultural experimental design. For more advanced applications and custom designs, consult the package documentation and additional vignettes.*
